---
title: "Guided Analyais"
author: "Eric J. Kort"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guided Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Identifying drugs targeting specific gene perturbations is facilitated by 
the `relinc` and `slinky` packages.  This document describes how to conduct 
such an analysis.

## Pre-requisites.

The relinc package is designed to expedite analysis by storing zscores 
on a redis server.  Do not open your redis server to the world.  If working 
on AWS, open port 6379 only to other servers within a specified security 
group.  Use similar strategies in other environments to keep your redis 
instance behind a firewall.  If you do not, you will find your redis server 
hacked within minutes.

The following tutorial was completed using an m5.4xlarge instance type 
($0.77/hour for on demand, presently $0.25/hour for spot pricing) running 
Ubuntu Server 20.04 LTS (HVM) (ami-0801628222e2e96d6), with a 100GB root volume. 

First we need to install some system libraries for R, and redis.

```{bash, eval=FALSE}
sudo -s
apt-get update
apt-get -y install build-essential
apt-get -y install gfortran
apt-get -y install libssl-dev
apt-get -y install libcurl4-openssl-dev
apt-get -y install libxml2-dev
apt-get -y install libtool-bin
apt-get -y install libhiredis-dev
apt-get -y install emacs-nox

sudo apt-get -y install redis-server

```

We can then install the latest R (at least, the latest relase from version 4)

```{bash, eval=FALSE}
# a little shell magic to get our Ubuntu release's "codename"
RELEASE=$(lsb_release -as 2>/dev/null | tail -n1)

echo "deb https://cloud.r-project.org/bin/linux/ubuntu ${RELEASE}-cran40/" >> /etc/apt/sources.list
gpg --keyserver keyserver.ubuntu.com --recv-keys 51716619E084DAB9
gpg -a --export 51716619E084DAB9 | sudo apt-key add -
apt-get update
apt-get -y install r-base

```

We are now ready to install libraries we will need to complete the analysis. 
From within an R session, run the following to check for required libraries 
and install any that are missing.

```{r, eval = FALSE}

req <- c(
    "R6",
    "doMC",
    "foreach",
    "parallel",
    "slinky",
    "redux",
    "flock",
    "dplyr",
    "devtools")

ix <- which(req %in% installed.packages())
if(length(ix < length(req))) {
  setRepositories(ind = c(1:5))
  install.packages(req[-ix])
}

if(!require(relinc))
  devtools::install_github("vanandelinstitute/relinc")

```

Make sure the redis-server is configured to store the snapshot somewhere with
enough room (the default location, `/var/lib/redis`, is fine if your root drive
has 50GB or so free), and to allow connections from outside (by commenting out
the "bind" line in the network section of redis.conf) if you are going to use
separate servers to do the analysis (see below).  Remember, "outside" means
"other servers in your security group or behind your firewall", not the entire
world.

Significance testing requires boostrapping large number of samples, and thus 
having many cores is desirable.  For example, we use 100,000 boostrapped 
samples for gene significance testing, which takes about 8 hours with 16 
cores. If you are working in the cloud, you may not want to keep such a beefy 
server running all the time. The recommended approach is to have your 
redis server on a smaller server (a couple of processors and 30GB of RAM is 
recommended) which you leave running until prolonged periods of downtime 
are anticipated, and then one or more larger multi-core servers in the same 
security group when you are actively selecting drugs.


The analysis also requires the LINCS L1000 data files (expression data and 
instance "info").  These need to be fetched and stored somewhere memorable 
that has enough room for them.  Once you have loaded the z-scores, you no 
longer need the expression data file.

```bash
mkdir data
cd data
wget ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE92nnn/GSE92742/suppl/GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx.gz
wget ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE92nnn/GSE92742/suppl/GSE92742_Broad_LINCS_inst_info.txt.gz
gunzip GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx.gz
```

## Calculating zscores

Once your redis server is up and running, you can configure the client and 
calculate and store robust zscores as follows.  The first command will prompt
you for details about the location of your redis server and data files.  You 
will also be prompted to save this information so you do not need to enter it 
each time.

```{r eval=FALSE}
library(relinc)
rl <- Relincr$new(interactive = TRUE)
rl$zscore_all_drugs()
rl$zscore_all_genes()

```

That will take quite a while--several hours or so, depending on how many cores
you have. Note that the you will experience diminishing returns at this step 
with lots of cores, since this requires calls to the clue.io api to identify 
appropriate control samples for each instance.  Relincr throttles these 
requests to to more than 1 every 2 seconds (even if using multiple cores). 
Since calculating the zscores also take a short bit of time, this is more 
efficient if you have multiple cores, but probably having more than 4-8 cores 
will not help any.  (This is not true when bootstrapping p-values later on--
for that task, the more parallelism the better) 

Although it takes a while, this only needs to be done once.  Once you have 
zscores calculated and loaded, you can use them forever assuming you hang 
on to the redis dump file--dump.rdb by default--so you can reload the data 
automatically in the event of a server reboot.  For this reason, you may 
want to configure redis to store the dump file somewhere non-ethereal (such 
as a dedicated EBS volume you create and mount for this purpose). If you do 
decide to do that, remember that you must both update the `dir` setting in 
redis.conf **and** add your dump file directory to 
`/etc/systemd/system/redis-server` with a line something like this (but 
with the actual path to the mount you created):

```{bash}
ReadWriteDirectories=/mnt/data/redis/dump/directory
```

## Gene scoring

The next step is to identify the gene signature for the gene perturbation of 
analysis.  First we score the genes using the rank-of-rank algorithm, as 
follows:

```{r eval=FALSE}
scores <- rl$ks_pert("LMNA", "trt_sh")
```

This will find all the instances where LMNA was treated with a perturbation of 
type "trt_sh" (i.e., knocked down with a short hairpin)
